Less Funding, More Lecturers, and Fewer Professors: \\ \vspace{0.1cm} \Large{Stagnating State Funding for Higher Education and its Effect on Faculty at Public Universities}

### 2023-12-20 text/paper.tex
Writing updates, Table A1 needs re-do.
Updates.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Title page
% Author
\author{Senan Hogan-Hennessy\thanks{
        For helpful comments I thank
        Levon Barseghyan,
        Francine Blau,
        Ryan Dycus,
        Ronald Ehrenberg,
        Michael Lovenheim,
        Jake Meyer,
        Douglas Miller,
        and Evan Riehl.
        I thank seminar participants at Cornell University (2022, 2023) for helpful discussion.} \\
    \vspace{0.1cm}
    Economics Department, Cornell University
}
% Title
\title{Less Funding, More Lecturers, and Fewer Professors: \\ \vspace{0.1cm}
    \Large{Stagnating State Funding for Higher Education and its Effect on Faculty at Public Universities}}
\date{This version: \today\thanks{
    This project's Github repository, with materials and anonymised data for replication, is available at 
    \url{https://github.com/shoganhennessy/state-funding-faculty}.
    Any comments or suggestions may be sent to me at \href{mailto:seh325@cornell.edu}{\nolinkurl{seh325@cornell.edu}}, or raised as an issue on the Github project.
    } \\ \vspace{0.1cm}
    \href{https://github.com/shoganhennessy/state-funding-faculty/blob/main/state-funding-faculty-2023.pdf}{Newest version available here.}
}
\rhead{\today}
\lhead{Less Funding, More Lecturers, and Fewer Professors.}
% Begin
\begin{document}
\maketitle
\thispagestyle{empty}
% Abstract
\begin{abstract}
    \noindent
    \input{sections/00-abstract.tex}
\end{abstract}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Starting the paper
\newpage
\setcounter{page}{1}
\doublespacing
% Introduction section
\noindent
\input{sections/01-introduction.tex}
% Data section
\input{sections/02-data.tex}
% Conceptual Framework
\input{sections/03-concepts.tex}
% Empirical methods
\input{sections/04-empirical.tex}
% Empirical results
\input{sections/05-results.tex}
% Dicussion
\input{sections/06-discussion.tex}
% Conclusion
\input{sections/07-conclusion.tex}

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Title page
% Author
\author{Senan Hogan-Hennessy\thanks{
        For helpful comments I thank
        Levon Barseghyan,
        Francine Blau,
        Ryan Dycus,
        Ronald Ehrenberg,
        Michael Lovenheim,
        Jake Meyer,
        Douglas Miller,
        and Evan Riehl.
        I thank seminar participants at Cornell University (2022, 2023) for helpful discussion.
        This project's repository, with all materials and anonymised data needed for replication, is available at 
        \url{https://github.com/shoganhennessy/state-funding-faculty}.
        Any comments or suggestions may be sent to me at \href{mailto:seh325@cornell.edu}{\nolinkurl{seh325@cornell.edu}}, or raised as an issue on the Github project.
    } \\
    \vspace{0.1cm}
    Economics Department, Cornell University\footnote{
        Address: Uris Hall \#447, Economics Department, Cornell University NY 14853 USA.
    }
}
% Title
\title{Less Funding, More Lecturers, and Fewer Professors: \\ \vspace{0.1cm}
    \Large{Stagnating State Funding for Higher Education and its Effect on Faculty at Public Universities}}
\date{This version: \today \\ \vspace{0.1cm}
    \href{https://github.com/shoganhennessy/state-funding-faculty/blob/main/state-funding-faculty-2023.pdf}{Newest version available here.}
}
\rhead{\today}
\lhead{Less Funding, More Lecturers, and Fewer Professors.}
% Begin
\begin{document}
\maketitle
\thispagestyle{empty}
% Abstract
\begin{abstract}
    \noindent
    \input{sections/00-abstract.tex}
\end{abstract}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Starting the paper
\newpage
\setcounter{page}{1}
\doublespacing
% Introduction section
\noindent
\input{sections/01-introduction.tex}
% Data section
\input{sections/02-data.tex}
% Conceptual Framework
\input{sections/03-concepts.tex}
% Empirical methods
\input{sections/04-empirical.tex}
% Empirical results
\input{sections/05-results.tex}
% Dicussion
\input{sections/06-discussion.tex}
% Conclusion
\input{sections/07-conclusion.tex}
% Rev of Econ Ed Statements.
\subsubsection*{Credit Author Statement}
As the only author on this paper, I am responsible for all contributions in this paper.
\subsubsection*{Declaration of Competing Interest}
This paper's author has no competing interests to declare.

### 2023-12-20 text/sections/02-data.tex
Updates.

States vary largely in how much they fund their public university systems, thanks in part to the state-wide budgeting process.
Planning for an annual budget begins two years ahead of the fiscal year, and the legislature votes to approve or reject the governor office's budget request a number of months before the fiscal year begins.\footnote{
    \cite{NBERw23736} present a full discussion of the decision-making process for state appropriations, drawing on administrative records originally analysed by \cite{parmley2009state}.
}
US state governments, by majority, are legally obligated to run a balanced-budget, so that yearly variation in tax revenues (caused by changing economic conditions or otherwise) necessarily affect state expenditures.
Public universities have relatively lower lobbying power than other funding beneificaries, so often bear the brunt of state government funding cuts \citep{delaney2011state}.\footnote{
    \cite{delaney2011state} fully describe the financial environment of state expenditures, and what makes spending on higher education an attractive area for state governments to expand funding during years of higher tax revenues, and retract funding in leaner years.
    An analysis of state expenditures for the years 1980-2004 (overlapping with the sample for this analysis) provides solid evidence for these trends, and \autoref{fig:funding} observes these same trends.
}
These features lead to yearly variation in state funding not seen in other revenues sources (such as federal funding), since US states differ in their support for higher education, and are subject to fiscal constraints.
Additionally, the number of higher education students in each state varies thanks to the size of each birth cohort.
For example, the birth cohort of 1971 was larger than for neighbouring years, leading to more student demand for limited public university places for students turning 18 in 1989 \citep{bound2007cohort}.
This leads to variation in the amount of funding per student across years, showing that per student state funding varies on multiple dimensions.

@@

There are large differences in the average number of professors per student between the private and public sector.
\autoref{fig:fte-perprof} shows that private universities start with a higher baseline of around 4.5 professors per hundred students, and exhibited yearly variation of less than 0.5 professor per hundred students over the thirty years.
Public universities start with 3.9 professors per hundred students, and this number falls to around 3., with the largest fall in the 2008-2011 time period.
We see a similar difference in baseline, and fall for the years 2008-2011 for assistant professors.
Private and public universities have similar numbers of associated and full professors before the year 2000, yet this number has fallen by over 20\% in the next 20 years only for public universities: in 2021 the mean public university has 6 fewer full professors per hundred students than the mean private university.
Over the same time period, we see the rise in use of non-tenure track instructor positions (referred to as lecturers from here on), who were employed at similar rates in both sectors in 1990, yet have been utilised by public universities at a higher rate since.

@@

Illinois funding for higher education has stagnated, and has experienced serious declines in the decade 2010--2021 --- similar in magnitude to the nation-wide decline over 1990--2021.
State funding among the seven Illinois public university campuses has fallen by over 50\% for the last decade, on both an absolute and per student basis (see \autoref{fig:illinois-funding}).
Similar to the figures for the rest of the country, there has been a corresponding substitution towards tuition revenue.

>>>>

States vary largely in how much they fund their public university systems, thanks in part to the state-wide budgeting process.
Planning for an annual budget begins two years ahead of the fiscal year, and the legislature votes to approve or reject the governor office's budget request a number of months before the fiscal year begins.\footnote{
    \cite{NBERw23736} present a full discussion of the decision-making process for state appropriations, drawing on administrative records originally analysed by \cite{parmley2009state}.
}
US state governments, by majority, are legally obligated to run a balanced-budget, so that yearly variation in tax revenues (caused by changing economic conditions or otherwise) necessarily affect state expenditures.
Public universities have relatively lower lobbying power than other funding beneficiaries, so often bear the brunt of state government funding cuts \citep{delaney2011state}.\footnote{
    \cite{delaney2011state} fully describe the financial environment of state expenditures, and what makes spending on higher education an attractive area for state governments to expand funding during years of higher tax revenues, and retract funding in leaner years.
    An analysis of state expenditures for the years 1980-2004 (overlapping with the sample for this analysis) provides solid evidence for these trends, and \autoref{fig:funding} observes these same trends.
}
These features lead to yearly variation in state funding not seen in other revenues sources (such as federal funding), since US states differ in their support for higher education, and are subject to fiscal constraints.
Additionally, the number of higher education students in each state varies thanks to the size of each birth cohort.
For example, the birth cohort of 1971 was larger than for neighbouring years, leading to more student demand for limited public university places for students turning 18 in 1989 \citep{bound2007cohort}.
This leads to variation in the amount of funding per student across years, showing that per student state funding varies on multiple dimensions.

@@

There are large differences in the average number of professors per student between the private and public sector.
\autoref{fig:fte-perprof} shows that private universities start with an average of 40 students for every one full professor, and little average trend until 2021.
Public universities start with 40 students per professor, and by 2023 there are 52.4 students professors for every full professor --- with the largest rise coming in the 2008-2011 time period.
The trend, and divergence between public and private universities, is similar for the number of assistant professors.
Private and public universities have similar numbers of associated and full professors before the year 2000, yet this number has fallen by over 20\% in the next 20 years only for public universities: in 2021 the mean public university has 6 fewer full professors per hundred students than the mean private university.
Over the same time period, we see the rise in use of non-tenure track instructor positions (collectively referred to as  lecturers), who were employed at similar rates in both sectors in 1990, yet have been utilised by public universities at a higher rate since.

@@

Illinois funding for higher education has stagnated, and has experienced serious declines in the decade 2010--2021 --- similar in magnitude to the nation-wide decline over 1990--2021.
State funding among the seven Illinois public university campuses has fallen by over 50\% for the last decade, on both an absolute and per-student basis (see \autoref{fig:illinois-funding}).
Similar to the figures for the rest of the country, there has been a corresponding substitution towards tuition revenue.

### 2023-12-20 text/sections/05-results.tex
Updates.

\subsection{First Stage Estimates}
The shift-share instrument has a strong first stage for universities' yearly state funding.
\autoref{tab:firststage-reg} presents results of the first-stage regression among IPEDS data, separately with and without a control for tuition revenue per student, plus institution and year fixed effects.
I use the same empirical strategy in the analysis of Illinois faculty, so  \autoref{tab:firststage-illinois} presents first-stage estimates among the IBHED data (which are very similar to those calculated in IPEDS data).
\autoref{tab:firststage-reg} Panel A shows estimates of a funding shock of \$1 per student in the state on state funding per student at the university, and Panel B the \% increase effects of a 1\% increase in the shock per student on state funding per student.
Column (1) shows that a funding shock of \$10 (per student in the entire state) is associated with \$11.76 less state funding per student at the university, with the corresponding -10\% funding shock leads to -9.77\%
less funding (column 1, panel B), with similar estimates with and without including fixed effects.
Column (2) shows estimates of the first-stage without including fixed effects, and gives less precise estimates for the funding shock, likely thanks to systematic differences in universities unaccounted for without fixed effects.
Columns (3) and (4) include the tuition revenue control (explained in \autoref{foot:control}) to exhibit estimates with the inclusion of this possible collider (or bad control).
Column (3) shows similar estimates to column (1) in both Panels A and B thanks to inclusion of fixed effects, so that the uncertainty in including tuition revenue as a possible bad control for the level of state funding does not matter thanks to the inclusion of fixed effects.
Column (1) represents the estimates for \autoref{eqn:firststage} with fixed effects, omitting the tuition revenue control, and is the preferred form that I proceed with.

The above shows that the first stage is strong among both IPEDS and IBHED data, and the argument in \autoref{sec:empirics} for exogenity comes from independence of the state funding mechanism to any indivdiaul public university campus.
\autoref{tab:firststage-balance} provides a balance test, showing that the shift-share instrument is correlated with the professor outcomes, but not with other sources of university funding.
These results, together with the case for exogeneity, show the funding shock instrument
strongly predicts university revenues in the first-stage estimation.

@@

\subsection{Reduced Hiring of New Professors}
Public universities substituted professors for more lecturers thanks to stagnating state funding, but there are little discernible effects on incumbent professors or lecturers.
This implies that the substitution arose by impacting the hiring of new professors to replace those retiring or leaving academia.
% \textbf{Expand this paragraph}.
I take recent data on the total faculty hired at US public universities over 2011--2021 (provided by \citealt{wapman2022quantifying,wapman2022zenodo}), to show that universities with lower state funding (per student) also hired fewer professors in the same time period --- shown in \autoref{fig:hiring-correlation}.
The shift-share IV model shows that a $10$\% cut in per student state funding leads to $13$\% fewer faculty hires per student across the decade 2011--2021.\footnote{
    This limited sample of data are a cross-section for count of professor hires 2011--2021, so that local projection estimates are not possible here.
    See \autoref{sec:appendix-hiring} for further details.
}
These results provide suggestive evidence that funding cuts caused faculty substitution by disrupting hiring of new professors at public universities.
This provides a coherent mechanism for how incumbent faculty (those who are already hired at the university) were not meaningfully impacted by funding cuts, yet their wider composition nonetheless changed when their universities experienced funding cuts.

>>>>

\subsection{First Stage Estimates}
The shift-share instrument has a strong first stage for universities' yearly state funding.
\autoref{tab:firststage-reg} presents results of the first-stage regression among IPEDS data, separately with and without a control for tuition revenue per student, plus institution and year fixed effects.
I use the same empirical strategy in the analysis of Illinois faculty, and \autoref{tab:firststage-illinois} presents first-stage estimates among the IBHED data ; results are similar to those calculated in IPEDS data.
\autoref{tab:firststage-reg} Panel A shows estimates of a state-wide funding shock of \$1 per student on state funding per student at the university, and Panel B the effects of a 1\% increase in the state-wide funding shock on state funding per student.
Column (1) shows that a state-wide funding shock of \$10 per student is associated with \$11.76 less state funding per student at the university; -10\% state-wide funding shock leads to -9.77\% state funding per student at the average university.
Estimates with and without including fixed effects.
Column (2) shows estimates of the first-stage without including fixed effects, and gives less precise estimates for the funding shock --- this may be due to systematic differences in universities, which are not accounted for without fixed effects.
Columns (3) and (4) include the tuition revenue as a control (explained in \autoref{foot:control}) to exhibit estimates with or without this control variable.
Column (3) shows similar estimates to column (1) in both Panels A and B thanks to inclusion of fixed effects, so that 
whether tuition revenue is a bad control or not does not affect empirical estimates.
Column (1) represents the estimates for \autoref{eqn:firststage} with fixed effects (not including tuition revenue as a control variable), and is the preferred form that I proceed with.

The first stage is strong in both IPEDS and IBHED data, and the argument in \autoref{sec:empirics} for exogeneity comes from independence of the state funding mechanism to any individual public university campus.
\autoref{tab:firststage-balance} provides a balance test, showing that the shift-share instrument is correlated with the professor outcomes, but not with other sources of university funding.
These results, together with the case for exogeneity, show the funding shock instrument
strongly predicts university revenues in the first-stage estimation.

@@

\subsection{Reduced Hiring of New Professors}
Public universities substituted professors for more lecturers thanks to stagnating state funding, but there are little discernible effects on incumbent professors or lecturers.
This implies that the substitution arose by impacting the hiring of new professors to replace those retiring or leaving academia.
% \textbf{Expand this paragraph, referencing the figure, and the OLS tables for the appendix}.
I take recent data on the total faculty hired at US public universities over 2011--2021 (provided by \citealt{wapman2022quantifying,wapman2022zenodo}), to show that universities with lower state funding (per student) also hired fewer professors in the same time period --- shown in \autoref{fig:hiring-correlation}.
The shift-share IV model shows that a $10$\% per-student state funding cut leads to $13$\% fewer faculty hires per student across the decade 2011--2021.\footnote{
    This limited sample of data are a cross-section for count of professor hires 2011--2021, so that local projection estimates are not possible here.
    See \autoref{sec:appendix-hiring} for further details.
}

These results provide suggestive evidence that funding cuts caused faculty substitution by disrupting hiring of new professors at public universities.
This provides a coherent mechanism for how incumbent faculty (those who are already hired at the university) were not meaningfully impacted by funding cuts, yet their wider composition nonetheless changed when their universities experienced funding cuts.

### 2023-12-20 text/sections/09-figures-tables.tex
Updates.

>>>>

\newpage
\begin{figure}[H]
    \centering
    \singlespacing
    \caption{State Funding and Total Professor Hiring at Public Universities, Total for 2011--2021.}
    \includegraphics[width=0.8\textwidth]{figures/hiring-correlation.png}
    \label{fig:hiring-correlation}
    \justify
    \footnotesize
    \textbf{Note}:
    This figure shows the correlation between state funding and the number of total number of professors hired at public universities between 2011--2021, and the line is an OLS line of best-fit.
    Data for the number of faculty hired at public universities are provided by \cite{wapman2022quantifying,wapman2022zenodo}, and funding data from IPEDS.
\end{figure}

### 2023-12-20 text/sections/10-appendix.tex
Updates.

These results were produced by integrating the total count of faculty hires for 2010--2021 for the top-ranked 180 US universities with a sum of the funding variables, and then estimating the models specified in \autoref{sec:iv-model-uni}.
There were no observable differences in the hiring rate of male vs female faculty.

\begin{figure}[h!]
    \centering
    \singlespacing
    \caption{State Funding and Faculty Hired at Public Universities, Total for 2011--2021.}
    \includegraphics[width=0.8\textwidth]{figures/hiring-correlation.png}
    \label{fig:hiring-correlation}
    \justify
    \footnotesize
    \textbf{Note}:
    This figure shows the correlation between state funding and the number of total number of faculty hired at public universities between 2011--2021, and the line is a line of best-fit.
    Data for the number of faculty hired at public universities are provided by \cite{wapman2022quantifying,wapman2022zenodo}, and funding data from IPEDS.
\end{figure}

\newpage
\begin{table}[h!]
    \singlespacing
    \centering
    \caption{OLS and 2SLS Estimates for University Faculty Hires, in Illinois 2011--2021.}

>>>>

These results were produced by integrating the total count of professor hires for 2010--2021 for the top-ranked 180 US universities with a sum of the funding variables, and then estimating the models specified in \autoref{sec:iv-model-uni}.
There were no observable differences in the hiring rate of male vs female faculty.

\begin{table}[h!]
    \singlespacing
    \centering
    \caption{OLS and 2SLS Estimates for University Faculty Hires, in Illinois 2011--2021.}
